---
title: "Introduction to {ggplot2}"
author: "Template: Ian Hussey; Content: Gian Wegmueller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

dir.create("plots")

```

# Dependencies

```{r}

library(readr)
library(ggplot2)
# install.packages("datasauRus")
library(datasauRus) 
library(scales)
library(dplyr)
library(tidyr)
library(plotrix) 
# devtools::install_github("matthewbjane/ThemePark")
library(ThemePark)
library(patchwork)
library(janitor)
library(knitr)
library(kableExtra)

```

# Why plot data?

Summary statistics aren't enough! Data with the same summary statistics can look vastly different when plotted.

```{r}

# M and SD
datasaurus_dozen |>
  group_by(dataset) |>
  summarize(mean_x = mean(x),
             sd_x = sd(x),
             mean_y = mean(y),
             sd_y = sd(y)) |>
  mutate_if(is.numeric, round_half_up, digits = 2) |>
  kable(align = 'r')|>
  kable_classic(full_width = FALSE)

# correlation
datasaurus_dozen |>
  group_by(dataset) |>
  summarize(correlation = cor(x, y)) |>
  mutate_if(is.numeric, round_half_up, digits = 2) |>
  kable(align = 'r') |>
  kable_classic(full_width = FALSE)

```

Always plot your data!

```{r fig.height=10, fig.width=10}

ggplot(datasaurus_dozen, aes(x = x, y = y)) +
  geom_point() +
  facet_wrap(~dataset) +
  theme_minimal()

```

# Structure of a ggplot

Whereas the pipes (`%>%` and `|>`) are used to create tidy data wrangling and analysis workflows, ggplot functions are added together with `+`.

Function calls are applied in order as layers. Changing the order functions are called can therefore change the appearance of the plot.

```{r}

# get data
data_processed <- read_csv("../data/processed/data_processed.csv")

data_after_exclusions <- data_processed |>
  filter(exclude_amp == "include" & 
           n_items == 3 & 
           gender %in% c("male", "female")) 

```

```{r fig.height=5, fig.width=6}

# data and aesthetics calls
plot_1 <- 
  ggplot(data = data_after_exclusions,
         aes(x = mean_self_report,
             y = amp_score,
             color = gender,
             shape = gender)) +
  # draw lines manually
  geom_vline(xintercept = 4, linetype = "dotted") +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  # draw geoms using the aesthetics (x, y, color and shape)
  ## points
  geom_point() +
  ## fit curves, in this case a linear model
  geom_smooth(method = "lm") +
  # adjust axis labels and ranges
  scale_x_continuous(name = "Explicit evaluation\n(Self-report)",
                     breaks = scales::breaks_pretty(n = 7)) +
  scale_y_continuous(name = "Implicit evaluation\n(AMP)") +
  # apply a theme
  theme_linedraw() + 
  # adjust elements of the theme
  labs(title = "Scatter plot with linear regression lines",
       color = "Gender",
       shape = "Gender") +
  # adjust the colors 
  scale_color_manual(values = c("female" = "#FF69B4",
                                "male" = "#6495ED"),
                     labels = c("female" = "Female",
                                "male" = "Male")) +
  # adjust the shapes
  scale_shape_manual(values = c("female" = 16, 
                                "male" = 17),
                     labels = c("female" = "Female",
                                "male" = "Male")) +
  # display specific x and y coordinates without dropping data points (nb using `limits` drops data points, coord_cartesian does not). Choose to plot just a subset of the data points, without changing the calculations!
  coord_cartesian(xlim = c(1, 7),
                  ylim = c(0, 1))

# display plot below chunk
plot_1

# save plot to disk as png
ggsave(plot = plot_1,
       filename = "plots/plot_1.pdf", 
       width = 6,
       height = 5)

```

Note that you can add additional function calls to objects later, e.g., overriding the previous theme_ call with a new one:

```{r}

plot_1 + theme_friends()

```

# Histogram using `geom_histogram()`

## Simple plot for self-reports

```{r}

ggplot(data = data_after_exclusions,
       aes(x = mean_self_report)) +
  geom_histogram()

```

## Slightly better plot for self-reports

```{r}

ggplot(data = data_after_exclusions,
       aes(x = mean_self_report)) +
  # more intelligent choices for the binwidth and boundary
  geom_histogram(binwidth = 1, boundary = 0.5) + # lowest possible answer is 1, so starting 0.5 creates more intuitive bins
  # labeling of the axis points
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), # seven breaks at reasonable intervals :)
                     limits = c(0.5, 7.5)) +
  theme_minimal()

```

## Exercise: Plot for gender

Create a similar plot for the gender variable in `data_processed` (ie before exclusions).

```{r}



```

## Exercise: Plot for AMP

Create a similar plot for the AMP scores in `data_after_exclusions`.

```{r}
mean_amp <- data_after_exclusions %>% 
  summarize(mean_amp = mean(amp_score)) %>% 
  pull(mean_amp)


ggplot(data = data_after_exclusions, 
       aes(x = amp_score)) +
  geom_histogram(binwidth = 0.1, boundary = -0.05) +
  scale_x_continuous(breaks = seq(0, 1, .1), 
                     name = "AMP Score") +
  scale_y_continuous(breaks = seq(0, 35, 5),
                     name = "Frequency") +
  geom_vline(xintercept = mean_amp, linetype = "dashed", colour = "green")
  theme_minimal()
  


```

- Exercise: How to add a dashed vertical line at the sample's mean AMP score?


# Density plot using `geom_density()`

## Simple plot for self-reports

```{r}

ggplot(data = data_after_exclusions,
       aes(x = mean_self_report)) +
  # more intelligent choices for the binwidth and boundary
  geom_density(adjust = 1, # the degree of smoothing can be adjusted here (above 1 gets closer to normal )
               color = "#FF69B4",
               fill = "darkblue", 
               alpha = 0.3) +
  # labeling of the axis points
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7),
                     limits = c(1, 7)) +
  theme_minimal()

```

## Exercise: Plot for AMP

Make a similar density plot for the AMP. 

- Add a theme. 
- Make the X axis breaks prettier. 
- Name both axis names more clearly.  

```{r}

ggplot(data = data_after_exclusions, 
      aes(x = amp_score)) +
  geom_density(adjust = 1, 
               color = "green",
               fill = "lightblue",
               alpha = 0.4) +
  scale_x_continuous(breaks = seq(0, 1, .1), 
                     name = "AMP Score") +
  theme_classic()


```


# Bar plot using `geom_col()`

Bar plots are bad and usually shouldn't be used. But they are sometimes unavoidable, so here's how to make them.

## Simple plot for AMP

```{r}

# create the summary values to be plotted
summary_amp <- data_after_exclusions %>%
  group_by(gender) %>%
  summarize(amp_mean = mean(amp_score),
            amp_se = plotrix::std.error(amp_score)) # apparently, no SE function exists in tidyverse or base R
  
# plot these values
ggplot(data = summary_amp, 
       aes(x = gender, 
           y = amp_mean)) +
  geom_col() +
  # geom_bar(stat = "identity") + # NB geom_col is equivalent to geom_bar when stat == "identity
  geom_linerange(aes(ymin = amp_mean - amp_se, 
                     ymax = amp_mean + amp_se)) 

```

## Slightly better plot for AMP

```{r}
  
ggplot(data = summary_amp, 
       aes(x = gender, 
           y = amp_mean)) +
  geom_col(fill = "#0b6623", # note that you can specify specific colors using hex codes or names
           color = "black", 
           width = 0.6) +
  geom_errorbar(aes(ymin = amp_mean - amp_se, 
                    ymax = amp_mean + amp_se), 
                width = 0.1, 
                color = "black") +
  labs(title = "Bar Plot of Gender with Standard Errors",
       x = "Gender",
       y = "Mean AMP score") +
  theme_linedraw() 

```

## Exercise: Plot for self-reports

Make a similar plot for the self-reports. 

- Exercise: How to capitalize 'Male' and 'Female' by wrangling the data before plotting?
- Use `coord_flip()` to swap the X and Y axes.

```{r}

# summary statistics
summary_selfreport <- data_after_exclusions %>%
  mutate(gender = stringr::str_to_title(gender)) %>% # str_to_title from stringr capitalizes the first letter of every string in a character vector (or character object in this case)
  group_by(gender) %>% 
  summarize(selfreport_mean = mean(mean_self_report),
            selfreport_se = plotrix::std.error(mean_self_report))

# bar plot
ggplot(data = summary_selfreport,
       aes(x = gender, 
           y = selfreport_mean)) +
  geom_col(fill = "#414487", # color from viridis color palette
           color = "black", # set color of frame
           width = 0.5) + # set width of bars
  geom_errorbar(aes(ymin = selfreport_mean - selfreport_se, # specify error bar aesthetics -> lower and upper limit
                    ymax = selfreport_mean + selfreport_se),
                width = 0.07, # set width of error bar 
                color = "black") +
  labs(title = "Bar Plot of Gender with Standard Errors",
       x = "Gender",
       y = "Mean Selfreport Score") +
  coord_flip() + # flip x and y axis of plot
  theme_minimal()

```



# Combining plots

```{r}

plot_all <- data_after_exclusions |>
  ggplot(aes(x = mean_self_report,
             y = amp_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("All")

plot_women <- data_after_exclusions |>
  filter(gender == "female") |>
  ggplot(aes(x = mean_self_report,
             y = amp_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Women")

plot_men <- data_after_exclusions |>
  filter(gender == "male") |>
  ggplot(aes(x = mean_self_report,
             y = amp_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Men")

# combine these plots with different arrangements
plot_women + plot_men

plot_women + plot_men + plot_layout(ncol = 1)

plot_all / (plot_women + plot_men)

```

# Faceting plots

Without repeating yourself, you can also make a plot for different subsets using `facet_wrap()` or `facet_grid()`

```{r fig.height=4, fig.width=8}

ggplot(data = data_after_exclusions,
         aes(x = mean_self_report,
             y = amp_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ gender) # create plots by gender and display plot for every value of gender next to each other

```

## Exercise

Create a plot that assesses the association between self report scores and AMP scores. By wrangling `data_processed` more prior to plotting, and using `facet_grid()`, compare a) men vs women and b) participants who are 30+ years old vs younger than 30.

Improve the appearance of the plot, including its text, colors, theme, etc.

```{r}

# data wrangling - adding a column age_category with values 30+ and sub 30 years
data_added_age_cat <- data_after_exclusions %>% 
  mutate(age_category = ifelse(age >= 30, "30+ years", "sub 30 years"))

# plots for association between self-reports and AMP scores by gender and age
ggplot(data = data_added_age_cat,
  aes(x = mean_self_report,
      y = amp_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(rows = vars(age_category),
             cols = vars(gender))

# needs a prettiness make-over :D



```

# Session info

```{r}

sessionInfo()

```



