---
title: "Reshaping data with {tidyr}' pivot_wider() and pivot_longer()"
author: "Template: Ian Hussey; Content: [student name]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Resources

See code and gifs [here](https://github.com/gadenbuie/tidyexplain?tab=readme-ov-file#pivot-wider-and-longer) which illustrate pivots (and indeed other tidyverse verbs).

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

#dir.create("plots")

```

# Dependencies

```{r}

library(dplyr)
library(tidyr)
library(tibble)
library(faux)
library(janitor)
library(ggplot2)
library(scales)
library(psych)
library(readr)
library(knitr)
library(kableExtra)

```

# Example

## Simulate data in wide format

```{r}

# set seed for reproducibility
set.seed(123)

# generate data 
data_wide <- 
  faux::rnorm_multi(n = 100,
                    vars = 5,
                    mu = 3,
                    sd = 1,
                    r = 0.5,
                    varnames = paste0("item_", 1:5),
                    empirical = FALSE) %>%
  rownames_to_column(var = "id")

# recode responses less than 1 or more than 5 to those values, then round scores to whole numbers
# note that {faux} has functions for doing this better
data_wide_likert <- data_wide %>%
  mutate(across(starts_with("item_"), ~ round_half_up(.x, digits = 0))) %>% # across() -> ask ChatGPT, .x is place holder for the thing you pass with the across() function, in this case all columns starting with "item_"
  mutate(across(starts_with("item_"), ~ ifelse(.x < 1, 1, ifelse(.x > 5, 5, .x)))) 

```

## Cronbach's alpha

Wide data like this is a) common and b) useful for calculating metrics like internal consistency.

```{r}

res_alpha <- data_wide_likert %>%
  select(starts_with("item_")) %>%
  psych::alpha()

cronbachs_alpha_estimate <- res_alpha$total$raw_alpha %>% 
  janitor::round_half_up(digits = 2)

```

Cronbach's $\alpha$ = `r cronbachs_alpha_estimate`

## Plot simulated data

```{r}

## Ian committed a coding sin here! He repeated code and changed only one thing (the item-number!)

ggplot(data_wide_likert, aes(x = item_1)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_2)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_3)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_4)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_5)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

```

- These plots repeat the mortal coding sin of repeating ourselves. If we reshaped the data to 'long' format we could use just one ggplot() call that includes facet_wrap(). 

## Reshape

Using `pivot_longer()`.

```{r}

data_long <- data_wide_likert %>%
  pivot_longer(cols = starts_with("item_"),
               names_to = "item",
               values_to = "response")

ggplot(data_long, aes(x = response)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw() +
  facet_wrap(~ item)

```

- What other ways could you specify this pivot_longer call's arguments?

### Calculate sum scores

```{r}

sum_scores <- data_long %>%
  group_by(id) %>%
  summarise(sum_score = sum(response))


ggplot(sum_scores, aes(x = sum_score)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  scale_x_continuous(breaks = breaks_pretty(n = 10)) +
  theme_linedraw()

```

## Convert this long data back to wide

Just to know how to do it.

```{r}

data_wide_again <- data_long %>%
  pivot_wider(names_from = item,
              values_from = response)

```

## Combine item and sum scores in one data frame

```{r}

data_item_and_sum_scores <- data_wide_again %>%
  left_join(sum_scores, by = "id")

# why joins are needed over bind_cols
# wrong <- bind_cols(data_wide_again, sum_scores) %>% select(-id) 
```

# Practice

Wrangle the demographics data included in this exercise more efficiently by reshaping it into wide format. Before, we used filter() to wrangle the age and gender data separately.

```{r}

# read data
data_demographics_raw <- read.csv("../data/raw/data_demographics_raw.csv")

# check if there are duplicates
data_demographics_raw |>
  count(subject) |>
  arrange(desc(n)) # yes, there is a duplicate entry for one subject! 2 seems to be the correct number of entries for each participant 

# select variables and remove duplicate entries
data_demographics_sub <- data_demographics_raw %>%
  select(unique_id = subject, item = trialcode, response) %>% # select variables of interest and renaming them
  distinct() # keep only unique rows from the data frame, since one participant has duplicate entries

# sanity check after removing duplicates
data_demographics_sub %>% 
  count(unique_id) %>% 
  arrange(desc(n)) 

# now all participants have just two entries!

# transform data to wide format
data_demographics_wide <- data_demographics_sub %>% 
  pivot_wider(names_from = item, # creates new columns from the values found in "item" (age and gender in this case)
              values_from = response) # takes values from the "response" column to put in cells

# cleaning the demographics data
data_demographics_clean <- data_demographics_wide %>% 
  mutate(age = stringr::str_remove_all(age, "[^\\d.]"),  # retains only digits and decimal points
         age = na_if(age, ""), # after retaining only digits, there are empty values -> make them NA
         age = as.numeric(age),
         gender = stringr::str_to_lower(gender), # all strings in gender to lower case
         gender = stringr::str_remove_all(gender, "[\\d.]"), # remove everything that is a digit between 0-9
         gender = na_if(gender, ""), # after removing digits, there are empty cells -> make them NA
         gender = case_when(gender == "woman" ~ "female", # if gender equals "woman", assign value "female"
                            gender == "yes" ~ NA_character_, # if gender equals "yes", assign NA
                            gender == "man" ~ "male", # if gender equals "man", assign value "male"
                            gender == "girl" ~ "female", # if gender equals "girl", assign value "female"
                            gender == "dude" ~ "male", # if gender equals "dude", assign value "male"
                            gender == "non binary" ~ "non-binary", # if gender equals "non binary", assing "non-binary"
                            TRUE ~ gender)) # anything else leave as it is (correct value "male" or "female")

  

```

# Session info

```{r}

sessionInfo()

```



