---
title: "Evaluations of positive and negative stimuli using the Affective Misattribution Procedure (AMP) and self-reports"
subtitle: "Data processing"
author: "Template: Ian Hussey; content: Gian Wegmueller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

```

# Dependencies

```{r}

library(dplyr)
library(tidyr)
library(readr)
library(janitor) # for clean_names()
library(stringr)

```

# Get data

```{r}

# demographics
data_demographics_raw_messy <- read_csv("../data/raw/data_demographics_raw_messy.csv", skip = 2) |> #relative path: two dots before the slash means "go up one folder" in the file hierarchy, relative to where this .rmd lives. 
  janitor::clean_names() #automatically cleans up names of variables

data_demographics_raw <- read_csv("../data/raw/data_demographics_raw.csv") |>
  janitor::clean_names()

# self report measure
data_selfreport_raw <- read_csv("../data/raw/data_selfreport_raw.csv") |>
  janitor::clean_names()

# affect attribution procedure
data_amp_raw <- read_csv("../data/raw/data_amp_raw.csv") |>
  janitor::clean_names()

```

# Demographics

Clean the demographics data so that summaries of age and gender can be made in the analysis.

```{r}

## pipe operator
# %>%  tidyverse pipe
# |> base R pipe


## FIRST ATTEMPTS AT DATA WRANGLING WITH VARIABLE AGE

# dat_age_gender <- data_demographics_raw |>  
#   select(subject, trialcode, response) |> # select the columns of interest
#   rename(unique_id = subject, item = trialcode) |>  # rename columns to more intuitive names
#   filter(item == "age") |>  # filter out all rows with "age" in column item
#   mutate(response = str_remove_all(response, "\\D"), #remove all non-numeric characters from the variable response (there is one value "old", because some participant felt funny..), ATTENTION: if a participant wrote their age in letters instead of numbers (e.g., twenty), we are loosing information. 
#          response = na_if(response, ""), #if response = "" (because "old" was removed and cell is now empty), assign NA
#          response = as.numeric(response)) #convert to numeric values


## TO DO FOR NEXT WEEK (07.10.2024)
# Do same kind of data wrangling for the gender variable, and commit and push to GitHub

## Create wide format data frame and clean up age and gender variables
dat_age_gender_wide <- data_demographics_raw |> 
  select(subject, trialcode, response) |> # select the columns of interest
  rename(unique_id = subject, item = trialcode) |> # rename columns to more intuitive names
  distinct(unique_id, item, response, .keep_all = TRUE) |> # there are duplicate entries for one unique_id, remove those duplicates, keep the first distinct row of values
  
  pivot_wider(names_from = item, # transform to wide with variable names coming from 'item'
              values_from = response) |> # and respective values from the variable 'response'; now the data frame contains the three columns unique_id, age, and gender
  
  mutate(age = str_remove_all(age, "\\D"), #remove all non-numeric characters from the variable age (there is one value "old", because some participant felt funny..), ATTENTION: if a participant wrote their age in letters instead of numbers (e.g., twenty), we are loosing information.
         age = na_if(age, ""), #if age = "" (because "old" was removed and cell is now empty), assign NA
         age = as.numeric(age)) |>  #convert to numeric values
  #what responses to gender do exist in the data? -> distinct(dat_age_gender_wide, gender)
  mutate(gender = case_when( #clean up responses to gender with case_when()
    gender %in% c("female", "FEMALE", "Female", "Woman") ~ "female", #if gender contains any of these versions of "female", assign the value "female"
    gender %in% c("Male", "male", "MALE") ~ "male", #if gender contains any of these versions of "male", assign the value "male"
    gender %in% c("non binary", "Non-Binary") ~ "non binary", #if gender contains any of these versions of "non binary", assign the value "non binary"
    TRUE ~ NA_character_ #assign NA to everything that does not match the expressions above (23 and yes in this case)
  ))
  


```

# Exclusions / data quality

## AMP

Create an exclusion variable `exclude_amp_performance` based on AMP performance: "exclude" if more than 10% of trials are < 100ms, or "include" if not.

Create an exclusion variable `exclude_amp_completeness` based on the AMP having the right number of trials.

```{r}

#data_amp_performance_criteria <- data_amp_raw 

```

# Self-reports

Create a mean score for each participant of the three self-report items.

```{r}

#data_selfreport_trial_level <- data_selfreport_raw 

```

# Affect Misattribution Procedure

Create an overall AMP bias score. Score each trial as having been evalauted as congruent with the prime or not. eg When trialcode == "prime_positive" & evaluative_response == 1, then record as 1, etc. AMP score is the proportion of congruent responses.

```{r}

#data_amp_score_congruence <- data_amp_raw 

```

# Combine 

Combine data frames together to create one wide-format data frame where each row represents a participant.

Flag all duplicates for exclusion.

```{r}

# data_processed_before_exclusions 

```

# Define master exclusion variable

```{r}

#data_processed <- data_processed_before_exclusions

```

# Write to disk

```{r}

# # in case this dir doesn't exist, create it
# dir.create("../data/processed/")
# 
# # save data to disk in that dir
# write_csv(data_processed, "../data/processed/data_processed.csv")

```

# Session info

```{r}

sessionInfo()

```


