---
title: "Plotting Assignment - "
author: "Gian Wegmueller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 


```

# Dependencies

```{r}

# update dependencies -> only include necessary packages

library(readr)
library(dplyr)
library(tidyr)
library(janitor)
library(stringr)
library(trackeR)
library(leaflet)
library(ggplot2)
library(ggtext)
library(hrbrthemes)
library(knitr)

```


# Data Wrangling


```{r Data exploration}
## Drug use data
drugs_raw <- read_csv("../data/drug_use_by_age.csv")


```

## Data Preparation Drug Use - Bad Plot

```{r Data prep}
drugs_comparison <- drugs_raw %>% 
  select(age, n, 
         starts_with(c("alcohol", "marijuana", "cocaine", "meth"))) %>% 
  filter(age %in% c("21", "22-23", "24-25", "26-29", "50-64")) %>% #choosing only legal drinking-age millenials for comparison here!
  mutate(cocaine_frequency = as.numeric(cocaine_frequency),
         meth_frequency = as.numeric(meth_frequency),
         generation = if_else(age == "50-64", "Boomers", "Millennials")) %>% 
  group_by(generation) %>% 
  summarize(booze_med_freq = janitor::round_half_up(weighted.mean(alcohol_frequency, n), 1),
            pot_med_freq = janitor::round_half_up(weighted.mean(marijuana_frequency, n), 1),
            coke_med_freq = janitor::round_half_up(weighted.mean(cocaine_frequency, n), 1),
            meth_med_freq = janitor::round_half_up(weighted.mean(meth_frequency, n), 1)) %>% 
  pivot_longer(cols = booze_med_freq:meth_med_freq,
               names_to = "drug",
               values_to = "median_freq")

```

## Data Preparation Lucerne Half-Marathon - Good Plot

```{r data prep lucerne half}

# loading the Garmin Connect activity data (requires readTCX-function from package trackeR)
lucerne_half_24 <- readTCX("../data/lucerne_half_2024.tcx")

# creating subset with variables of interest
lucerne_half_24_bpm <- lucerne_half_24 %>% 
  select(latitude, longitude, altitude, distance, heart_rate) %>% # selecting variables of interest
  mutate(distance_km = distance/1000, # creating distance_km variable by dividing 'distance' by 1000, because the latter is specified in metres
         heart_rate_zone = as.factor(case_when(heart_rate >= 137 & heart_rate <= 155 ~ "Zone 3", # creating heart_rate_zone variable according to data provided on Garmin Connect
                                               heart_rate >= 156 & heart_rate <= 175 ~ "Zone 4",
                                               heart_rate >= 176 ~ "Zone 5"))) %>% 
  relocate(latitude, longitude, distance_km, distance, altitude, heart_rate, heart_rate_zone) # reorder columns in data frame


### SLICED DATASET FOR INTERACTIVE MAP ###
# Due to the large number of observations in the complete dataset, the loop for rendering the interactive map takes up a lot of computational resources. Slicing the dataset in half by deleting every second row still retains more than enough data points to create a map without any significant loss of information

lucerne_half_24_sliced <- lucerne_half_24_bpm %>% 
  slice(seq(1, n(), 2))

```




# Plots
## Bad Plot - Median Use Frequency

```{r Visualization}

# Bad plot for visualizing median use frequency of 4 different drugs by generation (Boomers & Millennials)
drug_use_plot <- ggplot(data = drugs_comparison,
                 aes(x = median_freq,
                     y = drug,
                     colour = generation)) +
  
  geom_bin_2d(position = "stack", 
              binwidth = 5,
              fill = "#FF5F1F",
              show.legend = TRUE) +
  
  geom_line(mapping = aes(group = generation),
            linewidth = .2) +
  
  scale_colour_manual(values = c("#00BFFF", "#00827F")) + 
  
  scale_x_continuous(breaks = seq(0, 60, 2.3),
                     limits = c(0, 60)) +
  
  labs(title = "Median Use Frequency by Generation",
       x = "Median Use Frequency", 
       y = "",
       colour = "Generation") +
  
  theme_dark(base_family = "serif")
  

drug_use_plot
```


## Good Plot - Heart Rate Data from the Lucerne Half-Marathon 2024

The good plot visualizes my personal heart rate data from the Lucerne City Half-Marathon on October 27, 2024. I tracked the race with a Garmin Forerunner 255 running watch, which I have been using since July to collect training and overall fitness data. The Garmin Connect website allows you to download various types of information from all your activities, including GPS, pace, and heart rate data. According to the Garmin Connect log for the Lucerne Half-Marathon, my heart rate zones at the time of the race were: 

- Zone 1: 98 - 116 bpm
- Zone 2: 117 - 136 bpm
- Zone 3: 137 - 155 bpm
- Zone 4: 156 - 175 bpm
- Zone 5: $\geq$ 176 bpm



```{r data visualiziation lucerne half, fig.width=10, fig.height=5}

### Defining colour-blind friendly palettes using shades of usual heart rate zone colours.

# palette for colouring background zones in plot
pastel_palette <- c(
  "Zone 3" = "#CAF2C2",
  "Zone 4" = "#FFF8B8",
  "Zone 5" = "#FFD6C9"
)


# line colour palette for contrast
contrast_palette <- c(
  "Zone 3" = "#3B9E3B",
  "Zone 4" = "#E3A000",
  "Zone 5" = "#D3473F"
)

### Plotting

ggplot(data = lucerne_half_24_bpm, # define data
                          aes(x = distance_km, # distance on x-axis
                              y = heart_rate)) + # heart rate on y-axis
  
  # adding background for Zone 3
  annotate(geom = "rect", xmin = -0.2, xmax = max(lucerne_half_24_bpm$distance_km), 
           ymin = 140, ymax = 156, # defining limits on y-axis (156 not 155 as max for neat transition to next zone)
           fill = "#CAF2C2", # Zone 3 colour
           colour = NA, # no frame 
           alpha = 0.8) + # setting transparency
  
  # adding description of Zone 3
  annotate(geom = "text", 
           x = 21.3, y = 142, # setting position of text label
           label = "Zone 3 - Aerobic", 
           colour = "black", 
           fontface = "bold", # format to bold
           hjust = 1) + # right align
  
  # adding background for Zone 4
    annotate(geom = "rect", xmin = -0.2, xmax = max(lucerne_half_24_bpm$distance_km),
           ymin = 156, ymax = 175.5, # defining limits on y-axis (.5 value for cleaner transition to next zone)
           fill = "#FFF8B8", # Zone 4 colour
           colour = NA, # no frame
           alpha = 0.8) + # setting transparency
  
  # adding description of Zone 4
    annotate(geom = "text", 
             x = 21.3, y = 158, 
             label = "Zone 4 - Threshold", 
             colour = "black",
             fontface = "bold",
             hjust = 1) +
  
  # adding background for Zone 5
    annotate(geom = "rect", xmin = -0.2, xmax = max(lucerne_half_24_bpm$distance_km),
           ymin = 175.5, ymax = 193, # defining limits on y-axis
           fill = "#FFD6C9", # Zone 5 colour
           colour = NA, # no frame
           alpha = 0.8) + # setting transparency
  
  # adding description of Zone 5
    annotate(geom = "text", 
             x = 21.3, y = 177, 
             label = "Zone 5 - Max", 
             colour = "black",
             fontface = "bold",
             hjust = 1) +

  # adding heart rate line to plot
  geom_line(aes(x = distance_km, # setting aesthetics again
                y = heart_rate,
                colour = heart_rate_zone, # colour line according to heart rate zones
                group = 1), # makes sure just one line is added, not three separate lines for each HRZ
            linewidth = 1) + # setting line-thickness 
  
  # defining colours for geom_line
  scale_colour_manual(values = contrast_palette, # using contrast_palette defined above
                      guide = "none") + # no legend
  
  # formatting x-axis
  scale_x_continuous(breaks = seq(0, max(lucerne_half_24_bpm$distance_km), 1), # adding breaks for every km
                     limits = c(-0.2, max(lucerne_half_24_bpm$distance_km)), # setting limits of scale
                     expand = c(0, 0)) + # get rid of blank space between HRZ and axis
  
  # formatting y-axis 
  scale_y_continuous(breaks = seq(140, 190, 5), # adding breaks from 140 to 190 by steps of 5
                     limits = c(140, 193), # setting limits of scale
                     expand = c(0, 0)) + # get rid of blank space between HRZ and axis
  
  # adding text and titles
  labs(title = "Heart Rate Lucerne Half-Marathon 2024",
       subtitle = "A visualization of my personal heart rate data collected with a *Garmin Forerunner 255* running watch during the Lucerne City Half-Marathon on the 27th of October 2024. I crossed the finish line with a time of **1:39:25**, averaging **4:42 minutes** per kilometer.",
       caption = "<sup>*</sup>bpm = beats per minute") +
  
  xlab("Distance in km") +
  ylab("Heart Rate (bpm)<sup>*</sup>") +
  
  # setting theme
  theme_ipsum(base_family = "sans", # defining font
              base_size = 11, 
              subtitle_size = 12, # setting size for subtitle
              axis_title_size = 12, # setting size for axis titles
              axis_title_just = "lt", # left adjust axis titles
              plot_title_margin = 30, # setting margin for plot title
              subtitle_margin = 30, # setting margin for subtitle
              plot_margin = margin(15,15,15,15)) + # setting margins for plot
  
  # fine-tuning of theme
  theme(plot.title = element_textbox_simple(margin = margin(b = 15)), # defining margin for title textbox
        plot.subtitle = element_textbox_simple(margin = margin(b = 20)), # defining margin for subtitle textbox, automatic line breaks with element_textbox_simple()
        plot.caption = element_markdown(), # defining caption as markdown element (allows for superscript in text)
        panel.grid.major.y = element_blank(), # getting rid of major...
        panel.grid.minor.y = element_blank(), # ...and minor horizontal gridlines 
        panel.background = element_rect(colour = "darkgrey", size = 1), # adding a subtle frame to plot
        axis.title.x = element_text(margin = margin(10,0,0,0)), # setting margin for x-axis title
        axis.title.y = element_markdown(margin = margin(0,10,0,0), # setting margin for y-axis title, and define as markdown to allow for superscript notation
                                        hjust = 0.92)) # moving title on vertical dimension to align better with scale

```

### Interactive Map

In addition to the plot displaying the heart rate across the half-marathon distance, this interactive map displays the route of the Lucerne City Half-Marathon, conditionally coloured by heart rate zone. 

```{r}

### Create hear rate zone colour palette function using the contrast_palette
hrz_palette <- colorFactor(palette = contrast_palette, domain = lucerne_half_24_bpm$heart_rate_zone)

## interactive map to visualize route and heart rate zones
lucerne_half_map <- leaflet(lucerne_half_24_sliced) %>% 
  addProviderTiles(providers$CartoDB.Positron) # adding map layer

  # Loop through the data and add segments individually to colour them according to heart rate zone
for (i in 1:(nrow(lucerne_half_24_sliced) - 1)) {
  lucerne_half_map <- lucerne_half_map %>%   
  addPolylines(
      lng = c(lucerne_half_24_sliced$longitude[i], lucerne_half_24_sliced$longitude[i + 1]), # start of segment at the longitude of the ith observation, end of segment at the ith + 1 observation
      lat = c(lucerne_half_24_sliced$latitude[i], lucerne_half_24_sliced$latitude[i + 1]), # same here for latitude
      color = hrz_palette(lucerne_half_24_sliced$heart_rate_zone[i]),  # Color by heart rate zone of the ith obs.
      weight = 4, # stroke width in pixels
      opacity = 0.8 # stroke opacity
    )
}


lucerne_half_map <- lucerne_half_map %>% 
  
  # add start point marker to map
   addMarkers(lng = lucerne_half_24_sliced$longitude[1], lat = lucerne_half_24_sliced$latitude[1], # first GPS location
              popup = "Start", label = "Start", # add label to marker
             icon = icons(
               iconUrl = "https://upload.wikimedia.org/wikipedia/commons/c/ce/Green_dot_25.svg", # free to use icon URL
               iconWidth = 30, iconHeight = 30)) %>% # define icon size
  
  # add finish point marker to map
  addMarkers(lng = lucerne_half_24_sliced$longitude[2992], lat = lucerne_half_24_sliced$latitude[2992], # last GPS location
             popup = "Finish", label = "Finish", # add label to marker
             icon = icons(
               iconUrl = "https://upload.wikimedia.org/wikipedia/commons/7/7d/Japanese_crest_Maru_ni_Takanashi%28White_background%29.svg", # free to use icon URL
               iconWidth = 15, iconHeight = 15)) %>% # define icon size
  
  # add colour legend to map
  addLegend(position = "bottomleft",
            pal = hrz_palette,
            values = lucerne_half_24_sliced$heart_rate_zone,
            title = "Heart Rate Zone",
            opacity = 1) %>% 
  setView(lat = 47.0300000000000, lng = 8.318685121853448, zoom = 12.5) # setting default view of map


lucerne_half_map 


```





# Session info

```{r}

sessionInfo()

```
